Analysis of my running log
==========================
Some analysis of my running and racing logs. Last updated `r paste(Sys.Date())`.

R packages I use:
```{r LoadPackages, results='hide'}
require(RCurl, quietly=TRUE)
require(ggplot2, quietly=TRUE)
require(xtable, quietly=TRUE)
```


Weekly running volume
----------------------
A weekly running volume report from [Garmin Connect](http://connect.garmin.com/) was imported into Google Docs. The data was last imported on `r as.character(Sys.Date())`.

The following data processing operations will be done:
* Create variable `week` as a date variable using the character variable `Time.Period`
* Create variable `month` as a date variable using `week`
* Convert variable `Distance` to numeric
* Create variable `hours` as a numeric value of hours of running using the character variable `Time`

```{r ImportGarmin}
url <- getURL("https://docs.google.com/spreadsheet/pub?key=0ApyhYsT8Gi-EdEg3aDBNRGR6U2lvdU5idTJkY0dSYVE&single=true&gid=0&output=csv", cainfo = system.file("CurlSSL", "cacert.pem", package = "RCurl"))
classes <- c("character", "integer", rep("character", 8))
df <- read.csv(textConnection(url), header=TRUE, sep=",", colClasses=classes)
df$week <- as.Date(df$Time.Period, format="%m/%d/%Y")
df$month <- as.Date(paste("1", format(df$week, "%b %Y")), format="%d %b %Y")
df$Distance <- as.numeric(df$Distance)
df$Distance[is.na(df$Distance)] <- 0
Time <- df$Time
Time[Time == ":"] <- "00:00:00"
h <- rep(NULL, length(Time))
m <- rep(NULL, length(Time))
s <- rep(NULL, length(Time))
h <- as.numeric(sapply(strsplit(Time, ":"), "[[", 1))
m <- as.numeric(sapply(strsplit(Time, ":"), "[[", 2))
s <- as.numeric(sapply(strsplit(Time, ":"), "[[", 3))
is.incorrect <- s == 0
s[is.incorrect] <- m[is.incorrect]
m[is.incorrect] <- h[is.incorrect]
h[is.incorrect] <- 0
df$hours <- h + (m / 60) + (s / (60 * 60))
head(df)
```

The dataset includes weeks `r as.character(min(df$week))` to `r as.character(max(df$week))`.

Plot monthly mileage volume and time-on-feet. It is obvious that my training volume suffered greatly when I was teaching starting around March 2009 through June 2011. Also, Hazel was born in December 2009, which also affected my training volume.
```{r PlotMileage}
ggplot(df, aes(x=week, y=Distance)) +
  geom_smooth(size=1) +
  geom_point(alpha=1/3, size=3) +
  geom_hline(y=median(df$Distance)) +
  labs(title="Mileage volume", x="Date", y="Total distance (miles)")
```
```{r PlotTime}
ggplot(df, aes(x=week, y=hours)) +
  geom_smooth(size=1) +
  geom_point(alpha=1/3, size=3) +
  labs(title="Time-on-feet", x="Date", y="Total time-on-feet (hours)")
```


Races
-----
Here is some data on my marathon and ultramarathon race history. The data was taken from my [Marathon Maniacs race log](http://www.marathonmaniacsdb.com/Maniacs/MyMarathons.asp?ManiacId=369) and imported into Google Docs.
```{r RaceHistoryData}
url <- getURL("https://docs.google.com/spreadsheet/pub?key=0ApyhYsT8Gi-EdEh5azZuRGVYWnVkem0wck1WVEwyZ0E&single=true&gid=0&output=csv", cainfo=system.file("CurlSSL", "cacert.pem", package = "RCurl"))
df <- read.csv(textConnection(url), header=TRUE, sep=",")
df$date <- as.Date(df$date, format="%m/%d/%Y")
df$year <- as.integer(format(df$date, format="%Y"))
df$category <- rep(NA, length(df$distance))
df$category[df$distance == "Marathon"] <- "Marathon"
df$category[df$distance != "Marathon"] <- "Ultra"
df$km[df$distance == "Marathon"] <- 26.2 * 1.60934
df$distance <- as.character(df$distance)
is.km <- grepl("km", tolower(df$distance))
is.mi <- grepl("mi", tolower(df$distance))
list.km <- strsplit(df$distance[is.km], " ")
list.mi <- strsplit(df$distance[is.mi], " ")
df$km[is.km] <- as.numeric(sapply(list.km, "[[", 1))
df$km[is.mi] <- as.numeric(sapply(list.mi, "[[", 1)) * 1.60934
df <- subset(df, select=-c(distance))
head(df)
```

Plot the number of races by category by year. I started ultrarunning in 2008. By 2012, I switched from running primarily marathons to running ultramarathons.
```{r AggregateCategory, results='asis'}
df$count <- rep(1, nrow(df))
aggRaces <- aggregate(count ~ category + year, data=df, sum)
isMarathon <- aggRaces$category=="Marathon"
isUltra    <- aggRaces$category=="Ultra"
aggRaces$year <- factor(aggRaces$year)
aggRaces$cumsum[isMarathon] <- cumsum(aggRaces$count[isMarathon])
aggRaces$cumsum[isUltra   ] <- cumsum(aggRaces$count[isUltra   ])
print(xtable(aggRaces, digits=0), type="html")
incr <- 5
maxy <- incr * ceiling(sum(df$count) / incr)
ggplot(aggRaces, aes(x=year, weight=cumsum, fill=category)) +
  geom_bar() +
  labs(title="Cumulative number of races", x="Year", y="", fill="Distance") +
  scale_y_continuous(limits=c(0, maxy))
```

Plot the number of races by surface type by year. In 2008, I also ran my first trail race. By 2011, I switched primarily to trail racing.
```{r AggregateSurface, results='asis'}
df$count <- rep(1, nrow(df))
aggRaces <- aggregate(count ~ surface + year, data=df, sum)
isRoad  <- aggRaces$surface=="Road"
isTrail <- aggRaces$surface=="Trail"
aggRaces$year <- factor(aggRaces$year)
aggRaces$cumsum[isRoad ] <- cumsum(aggRaces$count[isRoad ])
aggRaces$cumsum[isTrail] <- cumsum(aggRaces$count[isTrail])
print(xtable(aggRaces, digits=0), type="html")
incr <- 5
maxy <- incr * ceiling(sum(df$count) / incr)
ggplot(aggRaces, aes(x=year, weight=cumsum, fill=surface)) +
  geom_bar() +
  labs(title="Cumulative number of races", x="Year", y="", fill="Surface") +
  scale_y_continuous(limits=c(0, maxy))
```
